Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-05 paper.Rmd
finished revising

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the complexity of the model and local explanation. For reference, the FIFA data contained 5000 observations of nine explanatory variables took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds total. PCA and statistics of the variables and attributions took 2.8 seconds. These runtimes were from a non-parallelized session on a modern laptop, but suffice to say that most of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP an excellent candidate to start. [Alternatively, the package __fastshap__ claims extremely low runtimes, attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting, @greenwell_fastshap_2020.]

@@

An implementation is provided in the open-source __R__ package __cheem__, available on CRAN at \url{https://CRAN.R-project.org/package=cheem}. Example data sets are provided, and you can upload your data after model fitting and computing the local explanations. In theory, this approach would work with any black box model, but the implementation currently only calculates tree SHAP for tree-based models supported by __treeshap__ [tree based models from __gbm__, __lightgbm__, __randomForest__, __ranger__, or __xgboost__  @greenwell_gbm_2020; @shi_lightgbm_2022; @liaw_classification_2002; @wright_ranger_2017; @chen_xgboost_2021, respectively]. Tree SHAP was selected because of its computational efficiency. The SHAP and oscillation explanations could be added with the use of the `DALEX::explain()` and would be an excellent direction to extend the work [@biecek_dalex_2018; @biecek_explanatory_2021].

<!--checklist asks for: Acknowledgments, Appendices, then References-->
## Package Infrastructure {#sec:infrastructure}

The above-described method and application are implemented as an open-source __R__ package, __cheem__ available on CRAN at \url{https://CRAN.R-project.org/package=cheem}. Preprocessing was facilitated with models created via __randomForest__ [@liaw_classification_2002] and explanations calculated with __treeshap__ [@kominsarczyk_treeshap_2021]. The application was made with __shiny__ [@chang_shiny_2021]. The tour visual is built with __spinifex__ [@spyrison_spinifex_2020]. Both views are created first with __ggplot2__ [@wickham_ggplot2_2016] and then rendered as interactive `html` widgets with __plotly__ [@sievert_interactive_2020]. __DALEX__ [@biecek_dalex_2018] and the free ebook, _Explanatory Model Analysis_ [@biecek_explanatory_2021] were a boon to understanding local explanations and how to apply them.

## Installation and Getting Started

>>>>

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the complexity of the model and local explanation. For reference, the FIFA data contained 5000 observations of nine explanatory variables took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds total. PCA and statistics of the variables and attributions took 2.8 seconds. These runtimes were from a non-parallelized session on a modern laptop, but suffice to say that most of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP an excellent candidate to start. Alternatively, the package __fastshap__ claims extremely low runtimes, attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting, @greenwell_fastshap_2020.

@@

<!--checklist asks for: Acknowledgments, Appendices, then References-->

# Package Infrastructure {#sec:infrastructure}

An implementation is provided in the open-source __R__ package __cheem__, available on CRAN at \url{https://CRAN.R-project.org/package=cheem}. Example data sets are provided, and you can upload your data after model fitting and computing the local explanations. The local explanations need to be pre-computed and uploaded. Examples show how to do this for tree SHAP values, using  __treeshap__ (tree based models from __gbm__, __lightgbm__, __randomForest__, __ranger__, or __xgboost__  @greenwell_gbm_2020; @shi_lightgbm_2022; @liaw_classification_2002; @wright_ranger_2017; @chen_xgboost_2021, respectively). The SHAP and oscillation explanations could be easily added using `DALEX::explain()` [@biecek_dalex_2018; @biecek_explanatory_2021].

The application was made with __shiny__ [@chang_shiny_2021]. The tour visual is built with __spinifex__ [@spyrison_spinifex_2020]. Both views are created first with __ggplot2__ [@wickham_ggplot2_2016] and then rendered as interactive `html` widgets with __plotly__ [@sievert_interactive_2020]. __DALEX__ [@biecek_dalex_2018] and the free ebook, _Explanatory Model Analysis_ [@biecek_explanatory_2021] were a boon to understanding local explanations and how to apply them.
